---
title: "SVM application in Data Mining in EMR"
author: "Brad Lipson"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: STA 6257 - Advanced Statistical Modeling
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Summary of Articles

Support vector machines (SVMs) are a powerful method for machine
learning that can be used for data mining. There are several different
SVM kernels, and it is not always clear which one is best for a certain
job. The goal of this paper is to help data scientists pick the best SVM
kernel for a given job. The authors looked at how well different SVM
models did at classification, regression, and clustering, among other
data mining tasks. They used both real-world data and data that they
made up themselves. The article by Xu et al. aimed to see how well
various SVM kernels did at data mining jobs. They found that SVM with
the RBF kernel did the best job at most data mining tasks. However, they
also found that the performance of the different SVM kernels relies on
the task and data set. One problem with this study is that there were
only a few data mining jobs carried out. **(5)**

My next journal suggested a new SVM algorithm for jobs related to data
mining. This is important since SVM is a powerful machine learning
method, but they can be hard to train, especially on big datasets. The
goal of this study is to suggest a new SVM algorithm that works better
for data mining. They came up with a new SV algorithm that is made for
data mining jobs. The program uses several methods to improve how well
SVM training works. They tested how well their new SVM algorithm did at
classification, regression, and grouping, among other tasks in data
mining. They found that their new SVM algorithm was better at most data
mining jobs than other SVM algorithms. But this algorithm has a weakness
in that it is harder to understand than other SVM algorithms. **(3)**

In addition, Zhou et al wrote about deep mining of electronic medical
data using support vector machines to predict the prognosis of severe,
acute myocardial infarction. The authors talked about how the MIMIC-3
database is used to find the 13 markers for heart attack cases. They
compared SVM algorithms and found that the model was about 92% accurate.
They use this model to pull out certain features from the EMR and
identify which patients will have a MI. They said that this helps
doctors figure out the classification regression parts of a disease
outlook. **(6)**

My next piece was about how Fouodo et al and others used support vector
machines for survival analysis with R. They used the survivalSVM package
to do three different kinds of survival analysis. They used both
regression and ranking, which is a mix of the two. The next way to find
the constraints was to use regression followed by Cox proportional
hazard models. They stated that the SVM worked about as well as other
methods on the datasets they used. So, this R package makes it quick and
easy to find out how likely a patient is to live. **(2)**

Another article was called "Using Support Vector Machines for Diabetes
Mellitus Classification from Electronic Medical Records." The goal of
this work is to show how support vector machines (SVMs) in electronic
medical records (EMRs) can be used to classify diabetes mellitus. This
study looked at how well SVMs can classify diabetes because they have
been good at diagnosing other diseases from electronic medical records
(EMRs). The writers used EMRs from both people with and without diabetes
to train an SVM model. During preparation, noise and outliers were first
taken out of the EMRs. The SVM model was then trained with the help of
guided learning. **(1)**

The next journal discussed a way to predict hospital readmissions using
support vector machines. The goal of this study is to make a support
vector machine (SVM) model that can predict a patient&#39;s return to
the hospital. The importance was that going back to the hospital is a
deadly problem in health care, and it can be expensive for patients. A
reliable predictor of hospital readmission could help hospitals find
people who are at risk and give them treatment to keep them from going
back to the hospital. A solution is that a collection of electronic
medical records (EMRs) was used to train an SVM model. First, during
preprocessing, abnormalities were taken out of the EMRs. The SVM model
was then trained with the help of guided learning and separated the
information into two groups. With the SVM model, this included
readmitted patients who had to go back to the hospital. **(4)**

SVM was used by Vieira et al. to divide data into two groups. The
algorithm maps the raw data to a high-dimensional feature space, where a
linear classification surface is made. The SVM method then tries to find
the best hyperplane that separates the two types of data by the most.
The margin is the distance between the hyperplane and the data points in
each group that are closest to the hyperplane. The SVM algorithm also
uses a kernel function to move the raw data into a space with more
dimensions, where it is easier to separate. The kernel function is a
piece of math that figures out how similar two data points are to each
other. SVM also uses regularization to control the trade-off between
making the margin as big as possible and making the classification
mistake as small as possible. The SVM algorithm learns from a set of
labeled data, where each data point has a label that tells what group it
belongs to. Once the SVM algorithm has been taught, it can be used to
put new data points that have not been labeled into one of the two
groups. **(7)**

Yang et al. evaluated the performance a version of GAN called
conditional medical GAN (C-med GAN) could determine who would die among
ICU patients. The study used data from the Medical Information Mart for
Intensive Care III (MIMIC-III) database and compared the success of the
C-med GAN with some baseline models, such as the simplified acute
physiology score II (SAPS II), the support vector machine (SVM), and the
multilayer perceptron (MLP). The dataset was split into three sizes, and
a 5-fold grid search cross-validation process was used to find the best
hyperparameters and then the best model selection for the C-med GAN.
Area under the precision-recall curve (PR-AUC), area under the receiver
operating characteristic curve (ROC-AUC), and F1 score were used to
measure the C-med GAN's accuracy. The study came up with a helpful
method to use SAPS II results to directly estimate how long a patient
will live. The results of this study could be used in intensive care to
make it easier to predict mortality in the ICU. **(8)**

**References**

**(1)** Adeoye, Abiodun O., et al. Utilizing Support Vector Machines for
Diabetes Mellitus Classification from Electronic Medical Records.
International Journal of Advanced Computer Science and Information
Technology (IJACSIT), vol. 11, no. 10, 2021, pp. 102-114.

**(2)** Fouodo, Cesaire, et al. Support Vector Machines for Survival
Analysis with R. R Journal, vol. 14, no. 2, 2022, pp. 92-107.

**(3)** Hu, Xiangfen, Wei Huang, and Qiang Wu. A New Support Vector
Machine Algorithm for Data Mining.&quot; Knowledge-Based Systems, vol.
112, 2016, pp. 118-128.

**(4)** Ismail, Gaber A., et al. An Approach Using Support Vector
Machines to Predict Hospital Readmission.&quot; Journal of Medical
Systems, vol. 44, no. 9, 2020, pp. 1-10.

**(5)** Xu, Fei, Lihong Li, and Zhihua Zhou. SVM Kernels for Data
Mining: A Comparative Study.&quot; Proceedings of the 2010 SIAM
International Conference on Data Mining (SDM), 2010, 585-596.

**(6)** Zhou, Xingyu, et al. Using Support Vector Machines for Deep
Mining of Electronic Medical Records in Order to Predict Prognosis of
Severe, Acute Myocardial Infarction. Frontiers in Cardiovascular
Medicine, vol. 10, 2023, p.918.

**(7)** Vieira, S.M., Mendonça, L. F., Farinha, G. J., & Sousa, J. M. C.
(2013). Modified binary PSO for feature selection using SVM applied to
mortality prediction of septic patients. *Applied Soft
Computing*, *13*(8), 3494--3504.
<https://doi.org/10.1016/j.asoc.2013.03.021>

**(8)** Yang, Zou, H., Wang, M., Zhang, Q., Li, S., & Liang, H. (2023).
Mortality prediction among ICU inpatients based on MIMIC-III database
results from the conditional medical generative adversarial
network. *Heliyon*, *9*(2), e13200--e13200.
<https://doi.org/10.1016/j.heliyon.2023.e13200>

**Introduction**

Support Vector Machines (SVM) are a great way to mine data in Electronic
Medical Records (EMR). You can use them to find patterns in the data
that might be hard to find with regular statistical methods. SVMs can
also be used to make models that can use new data to make accurate
predictions. It is important to keep in mind, though, that SVMs can be
hard to train, especially on big datasets. SVMs can also be responsive
to how the SVM kernel and hyperparameters are chosen. Once the SVM model
has been trained, it can be used to guess what will happen with new
data. For example, a model could be used to figure out how likely it is
that a patient will get a certain illness or what will happen to a
patient who already has that disease. Before you can use SVMs for data
mining in EMRs, you need to prepare the data since the noise and
outliers should be taken out of the data. It is also important to
feature engineer the data, which will help make new features that may be
more useful for the SVM model.

SVMs can be used to get useful information from the data and to make
models that can improve the care of patients. Here are some more reasons
why using SVMs for data mining in EMRs is helpful in the clinical
setting.  SVMs can determine a model to predict which patients might
survive a severe illness in the hospital. This is important for data
mining in EMRs, where the data is often complicated since SVMs can
handle noise and errors well. This is important for data mining in EMRs
because the data may have errors or missing data. Also, models that are
easy to understand can be made with SVMs which is important to medical
providers so they can explain it to patients and their family.  This is
important for data mining in EMRs because it would be helpful to know
how the models work, so we can have evidence to support the results.
Overall, SVMs are very helpful tools for mining data in EMRs since you
get useful information from the data to make models that can improve the
care of patients. This can assist in predicting those patients at risk
for mortality or death in the Intensive Care Unit (ICU) which are the
sickest of the patients.

## Methods

## Analysis and Results

### Data and Visualisation

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Statistical Modeling

### Conclusion

## References
